Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-03-05 paper/paper.tex
pass on Section 4.3

\textbf{Implementation details.} We implemented function
\emph{isPlausible} by 1)~updating Suricata to only monitor the rule
that is passed as parameter\Mar{isto realmente eh necessario? se sim,
  explique. me pareceu que o que vc. precisa ate aqui jah estaria no
  log do \suri}, 2)~replaying the negative benign traffic (see
Section~\ref{sec:dataset-benign}), and 3)~reading the Suricata logs to
check if any benign message was captured by the rule. It is worth
noting that the benign traffic is stored in \emph{pcap}
files~\cite{pcap} and Suricata can process those files very
efficiently compared to analyzing network messages directly. Instead
of sending traffic to a machine with a deployed server (\eg{}, web
server) that \suri\ would monitor, our setup calls \suri\ directly
through an API. \suri\ is invoked multiple times in the process of
minimizing the rule.  \textbf{Limitations.} It is worth noting that
the performance of \tname{} depends on the quality of inputs. For the
positive malicious traffic, \tname{} only requires one sample (as
input to step one), but can benefit of several samples (see
line~\ref{filtering} of Algorithm~\ref{algo1}). The same applies to
the negative traffic. However, in this case, \tname{} can use a large
variety of public data sets available online (see
Section~\ref{sec:objectanalysis-benign}).

@@

\tname{} uses a composition of heurisitic functions to sort the rules
produced in the previous step. Each of these functions uses a
different criterion to estimate how similar a created rule is to those
from public rulesets~\cite{emerging-threats-open}. The
intuition is that rules that look more similar to existing trusted
rules rank higher and could catch the attention of a system
administrator. The score of a rule is the weighted average across the
values produced by each heuristic function. The codomain of each
function ranges on the 0-1 interval. Hence, the score of a rule also
ranges on the 0-1 interval. We describe in the following each of these
heuristic functions.

@@

\subsubsection{Size of a rule}
\label{sec:heuristic-size-of-rule}
One element that \tname{} uses to measure similarity between a
synthesized and a human-created rule is the size of the synthetic
rule, \ie{}, the number of options it contains. Informally, rules with
popular sizes score high whereas unpopular rules score
low. Figure~\ref{fig:distribution-rule-size} shows the histogram of
rule sizes for our rule data set. The x-axis shows the number of
options included in a rule (up to 10) while the y-axis shows the
number of rules with the corresponding number of options. Note that
there are 11 rules with no options. These rules only use the header to
signal a traffic as malicious. Also note that most rules have 2 to 6
options. \tname{} values rules in that size range higher compared to
other sizes. We calculate the contribution of size of a given rule
$r$ with the following formula $\mathit{rules\_per\_size}[\mathit{len(r)}]/\mathit{max(rules\_per\_size)}$, where $\mathit{len(r)}$ denotes the number of options of a rule $r$,
$\mathit{rules\_per\_size}$ encodes the distribution of sizes as a
dictionary mapping sizes to number of associated rules, and
$\mathit{max(rules\_per\_size)}$ returns the maximum number of rules
for any dictionary entry as to normalize the measurement in the 0-1
range. Consequently, the contribution of size is proportional to its
popularity. The contribution of this heuristic would be 0.684
(=3,662/5,353) to a rule with three options and 0.028 (=154/5,353) to
a rule with nine options.

@@

@@

@@

          \addplot coordinates {(0,11) (1,1301) (2,4910) (3,3662) (4,3250) (5,5353) (6,2911) (7, 1337) (8, 1246) (9,763) (10, 495)};

@@

    \caption{\label{fig:distribution-rule-size}Rule sizes.}

@@

\begin{subfigure}{.2\textwidth}
  \caption{\label{fig:distribution-options}Options.}

@@

\subsubsection{Number of content options.} We observed
that \percContentOptions\ of the rules from the ruleset we are
analyzing contain content options, but the distribution of those
contents per rule is not uniform. Figure~\ref{fig:distribution-contents} shows the
corresponding distribution. Given the importance of this kind of
option to the definition of rules, we defined a heuristic function
focused on the typical number of content options in a rule. We
calculate the contribution of this function as described on previous
Section~\ref{sec:heuristic-size-of-rule}.

@@

\subsubsection{Popularity of options.}

\Mar{--------------------------------------------------------- estou aqui}

\Gui{A quinta heuristica que utilizamos
  considera a popularidade das opcoes. Primeiro pegamos a quantidade
  de vezes que uma opcao eh utilizada (sem considerar repeticoes numa
  unica regra). Dividimos o valor encontrado para cada opcao pelo
  maior valor encontrado, encontrando sempre valores de no maximo
  1. Entao na regra nos somamos os valores de cada opcao, dividido
  pelo numero de opcoes, chegando a um numero de no maximo 1, que
  representa a popularidade das opcoes na regra.
\[\frac{\sum_{i=1}^{N}\frac{\mathit{option\_frequency[option_i]}}{\mathit{max(option\_frequency)}}}{N}\]}

@@

\begin{table*}[t!]
  \small
  \caption{\label{table:attacks}List of attacks. Non-http attacks
    appear at the top; http attacks appear at the bottom.}
  \centering
  \begin{tabular}{llllp{10cm}}
    \toprule
    \multicolumn{1}{c}{Name} &
    \multicolumn{1}{c}{Source} &
    \multicolumn{1}{c}{Type} &
    \multicolumn{1}{c}{Protocol} &
    \multicolumn{1}{c}{Groud Truth (relevant/checked options)} \\
    \midrule
    Black Nurse & \cite{pcap-attacks} & DoS & ICMP & \textbf{icode}:3; \textbf{itype}:3; \textbf{threshold}: type both, track by\_dst, count 250, seconds 1;\\    
    Ping Scan & \cite{netmap} & AREC & ICMP & \textbf{dsize}:0; \textbf{itype}:8; \\
    SYN Flood & \cite{hping3} & DoS & TCP & \textbf{flags}: S,12;
    \textbf{threshold}: type both, track by\_dst, count 5000, seconds 5;\\
    UDP Teardrop & \Fix{WireShark} & DoS & UDP & \textbf{fragbits}:M; \textbf{id}:124; \\
    %%UDP Flood & MazeBolt & DoS & UDP & \textbf{fragbits}:M; \textbf{threshold}: type both, track by\_dst, count 5000, seconds 5; \\
    \midrule
    %%Apache Struts 2 & \cite{wrccdc} & RCE & HTTP & \textbf{content}:"POST"; http\_method; \textbf{content}:"java.lang.ProcessBuilder"; nocase; http\_client\_body; fast\_pattern; \textbf{content}:"/struts2-rest-showcase/orders/3"; http\_uri; \\    
    ColdFusion \MyComment{PESC} & \cite{nikto} & PESC & HTTP \MyComment{& Privilege Escalation} & \textbf{content}:"GET"; http\_method; nocase; \textbf{content}:"/CFIDE/administrator"; http\_uri; nocase; \\
    iisadmin access\MyComment{web-application-attack} & \cite{nikto} & RCE & HTTP \MyComment{& Adicionar} & \textbf{content}:"/iisadmin"; nocase; http\_uri; \\
    JBoss JMX\MyComment{ Console Beanshell Exploit} & \cite{nikto} & RCE & HTTP \MyComment{& Remote Code Execution} & \textbf{content}:"/HtmlAdaptor"; nocase; http\_uri; \textbf{content}:"action=inspect"; nocase; http\_uri; \textbf{content}:"bean"; nocase; http\_uri; \textbf{content}:"name="; http\_uri; \\
    Script tag in URI\MyComment{Possible Cross Site Scripting Attempt} & \cite{nikto} & RCE & HTTP \MyComment{& Adicionar} & \textbf{content}:"</script>"; nocase; http\_uri; \\
    WeBid LFI \MyComment{Local File Inclusion} & \cite{nikto} & LFI & HTTP & \textbf{content}:"GET "; depth:4; \textbf{content}:"/cron.php?"; http\_uri; nocase; \textbf{content}:"include\_path="; http\_uri; nocase; \textbf{content}:"../"; \\
    %%WordPress Login & \cite{wrccdc} & PVE & HTTP & \textbf{content}:"log="; http\_client\_body; \textbf{content}:"\&pwd="; http\_client\_body; \textbf{content}:"\&wp-submit="; http\_client\_body; \\
    .htaccess access\MyComment{attempted-recon} & \cite{nikto} & AREC & HTTP \MyComment{& Adicionar} & \textbf{content}:".htaccess"; nocase; http\_uri; \\
    .idq access\MyComment{EXPLOIT ISAPI} & \cite{nikto} & RCE & HTTP \MyComment{& Adicionar} & \textbf{content}:".idq"; nocase; http\_uri; \\
    /system32/ in URI\MyComment{Possible Protected Directory Access Attempt} & \cite{nikto} & AREC & HTTP \MyComment{& Adicionar} & \textbf{content}:"/system32/"; nocase; http\_uri; \\
    Basebuilder LFI & \cite{nikto} & LFI & HTTP & \textbf{content}:"GET"; http\_method; \textbf{content}:"/main.inc.php?"; nocase; http\_uri; \textbf{content}:"mj\_config[src\_path]="; nocase; http\_uri; \\
    Tomcat server snoop access & \cite{nikto} & AREC & HTTP & \textbf{content}:"/jsp/snp/"; http\_uri; \textbf{content}:".snp"; http\_uri;\\  
    \bottomrule
  \end{tabular}
\end{table*}

>>>>

\textbf{Implementation details.} We implemented function
\emph{isPlausible} by 1)~updating Suricata to only monitor the rule
that is passed as parameter\Mar{isto realmente eh necessario? se sim,
  explique. me pareceu que o que vc. precisa ate aqui jah estaria no
  log do \suri}, 2)~replaying the negative benign traffic (see
Section~\ref{sec:dataset-benign}), and 3)~reading the Suricata logs to
check if any benign message was captured by the rule. \Mar{discuss
  filter function.} It is worth
noting that the benign traffic is stored in \emph{pcap}
files~\cite{pcap} and Suricata can process those files very
efficiently compared to analyzing network messages directly. Instead
of sending traffic to a machine with a deployed server (\eg{}, web
server) that \suri\ would monitor, our setup calls \suri\ directly
through an API. \suri\ is invoked multiple times in the process of
minimizing the rule.  \textbf{Limitations.} It is worth noting that
the performance of \tname{} depends on the quality of inputs. For the
positive malicious traffic, \tname{} only requires one sample (as
input to step one), but can benefit of several samples (see
line~\ref{filtering} of Algorithm~\ref{algo1}). The same applies to
the negative traffic. However, in this case, \tname{} can use a large
variety of public data sets available online (see
Section~\ref{sec:objectanalysis-benign}).

@@

\tname{} uses a composition of heurisitic functions to sort the rules
produced in the previous step. Each of these functions uses a
different criterion to estimate how similar a created rule is to those
from public rulesets~\cite{emerging-threats-open}. The intuition is
that rules that look more similar to existing trusted rules are more
likely to be useful. The aggregated score of a rule is the weighted
average across the values produced by each heuristic function. The
codomain of each function ranges on the 0-1 interval. Hence, the score
of a rule also ranges on the 0-1 interval. We describe in the
following each of these heuristic functions.

@@

%% Note that there are 11 rules with no
%% options. These rules only use the header to signal a traffic as
%% malicious. Also note that most rules have 2 to 6 options.

\subsubsection{Popularity of options.}
One heuristic function that \tname{} uses to measure similarity
between a synthesized rule and a human-created rule is the popularity
of the options used in those rules. According to this heuristic, rules
containing popular options score higher compared to those containing
unpopular rules. Figure~\ref{fig:distribution-options} shows the
histogram of rule options for the ruleset we
used~\cite{emerging-threats-open}. The x-axis shows the options used
in rules (the 10 most common) and the y-axis shows the number of rules
using the corresponding option. Note that the options
\CodeIn{content}, \CodeIn{flow}, and \CodeIn{prce} are very
popular. \tname\ uses the following formula to compute the
contribution of option popularity to the score of a rule $r$:

\vspace{-2ex}
\[\frac{\sum_{i=1}^{N}\frac{\mathit{option\_frequency[option_i]}}{\mathit{max(option\_frequency)}}}{N}\]
\vspace{-2ex}

\noindent
The term $\mathit{option_i}$ denotes the $i$th option in the bitvector
encoding $r$. The term $\mathit{option\_frequency}$ encodes the
distribution of option popularity as a dictionary mapping
numbers---indicating popularity---to corresponding
options. Consequently, the term $\mathit{option\_frequency[option_i]}$
denotes the number of rules, from our dataset, containing
$\mathit{option\_i}$. The term $\mathit{max(option\_frequency)}$
refers to the maximum number of rules for any dictionary entry (to
normalize the measurement in the 0-1 range), and $N$ is the number of
options in $r$. To illustrate, the contribution of this function to
the score of a rule would be $0.26$ ($=(10181+3049)/(2*25699)$) for a
rule with options \CodeIn{prce} and \CodeIn{flowbits} and $0.01$
($=(155+108)/(2*25699)$) for a rule with options \CodeIn{itype} and
\CodeIn{icode}.

%% \Gui{A quinta heuristica que utilizamos
%%   considera a popularidade das opcoes. Primeiro pegamos a quantidade
%%   de vezes que uma opcao eh utilizada (sem considerar repeticoes numa
%%   unica regra). Dividimos o valor encontrado para cada opcao pelo
%%   maior valor encontrado, encontrando sempre valores de no maximo
%%   1. Entao na regra nos somamos os valores de cada opcao, dividido
%%   pelo numero de opcoes, chegando a um numero de no maximo 1, que
%%   representa a popularidade das opcoes na regra.}

%% \subsubsection{Size of a rule}
%% \label{sec:heuristic-size-of-rule}
%% \tname{} values rules in that size range higher compared to
%% other sizes.

%% We calculate the contribution of size of a given rule
%% $r$ with the  $\mathit{rules\_per\_size}[\mathit{len(r)}]/\mathit{max(rules\_per\_size)}$, where $\mathit{len(r)}$ denotes the number of options of a rule $r$,
%% $\mathit{rules\_per\_size}$ encodes the distribution of sizes as a
%% dictionary mapping sizes to number of associated rules, and
%% $\mathit{max(rules\_per\_size)}$ returns the maximum number of rules
%% for any dictionary entry as to normalize the measurement in the 0-1
%% range. 
%% Consequently, the contribution of size is proportional to its
%% popularity. The contribution of this heuristic would be 0.684
%% (=3,662/5,353) to a rule with three options and 0.028 (=154/5,353) to
%% a rule with nine options.

@@

%%   \begin{subfigure}{.25\textwidth}
%%     \centering
%%     \scalebox{0.8}{
%%       \begin{tikzpicture}
%%         \begin{axis}[
%%           bar width=4pt,
%%           scaled ticks=true,
%%           ymajorticks=false,
%%           ybar stacked,
%%           enlargelimits=0.08,
%% %          enlargelimits=false,
%% %          axis equal,
%%           clip=false,
%%           legend style={at={(0.5,-0.15)},anchor=north,legend columns=-1},          
%%           xtick=data,
%% %          xticklabel style={rotate=45},          
%%           nodes near coords,
%%           every node near coord/.append style={font=\scriptsize},
%%           xticklabel style={font=\footnotesize},          
%%           nodes near coords align={vertical},
%%           symbolic x coords={0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10},
%%           ]
%%           \addplot coordinates {(0,11) (1,1301) (2,4910) (3,3662) (4,3250) (5,5353) (6,2911) (7, 1337) (8, 1246) (9,763) (10, 495)};
%%         \end{axis}
%%       \end{tikzpicture}
%%     }
%%     \vspace{-1ex}
%%     \caption{\label{fig:distribution-rule-size}Rule sizes.}
%%   \end{subfigure}%

@@

          xticklabel style={rotate=30},

@@

@@

  \hspace{4ex}
  \caption{\label{fig:distribution-options}Option popularity.}

@@

@@

\Mar{------------------------------------------ estou aqui}
\subsubsection{Number of content options.} We observed
that \percContentOptions\ of the rules from the ruleset we are
analyzing contain content options, but the distribution of those
contents per rule is not uniform. Figure~\ref{fig:distribution-contents} shows the
corresponding distribution. Given the importance of this kind of
option to the definition of rules, we defined a heuristic function
focused on the typical number of content options in a rule. We
calculate the contribution of this function as described on previous
Section~\ref{sec:heuristic-size-of-rule}.

@@

\subsubsection{Optimization}

@@

\setlength{\tabcolsep}{5pt}
\begin{table*}[t!]
  \small
  \caption{\label{table:attacks}List of attacks. Non-http attacks
    appear at the top; http attacks appear at the bottom.}
  \centering
  \begin{tabular}{llllp{8cm}}
    \toprule
    \multicolumn{1}{c}{Name} &
    \multicolumn{1}{c}{Source} &
    \multicolumn{1}{c}{Type} &
    \multicolumn{1}{c}{Protocol} &
    \multicolumn{1}{c}{Groud Truth (relevant/checked options)} \\
    \midrule
    Black Nurse & \cite{pcap-attacks} & DoS & ICMP & \textbf{icode}:3; \textbf{itype}:3; \textbf{threshold}: type both, track by\_dst, count 250, seconds 1;\\    
    Ping Scan & \cite{netmap} & AREC & ICMP & \textbf{dsize}:0; \textbf{itype}:8; \\
    SYN Flood & \cite{hping3} & DoS & TCP & \textbf{flags}: S,12;
    \textbf{threshold}: type both, track by\_dst, count 5000, seconds 5;\\
    UDP Teardrop & \Fix{WireShark} & DoS & UDP & \textbf{fragbits}:M; \textbf{id}:124; \\
    %%UDP Flood & MazeBolt & DoS & UDP & \textbf{fragbits}:M; \textbf{threshold}: type both, track by\_dst, count 5000, seconds 5; \\
    \midrule
    %%Apache Struts 2 & \cite{wrccdc} & RCE & HTTP & \textbf{content}:"POST"; http\_method; \textbf{content}:"java.lang.ProcessBuilder"; nocase; http\_client\_body; fast\_pattern; \textbf{content}:"/struts2-rest-showcase/orders/3"; http\_uri; \\    
    ColdFusion \MyComment{PESC} & \cite{nikto} & PESC & HTTP \MyComment{& Privilege Escalation} & \textbf{content}:"GET"; http\_method; nocase; \textbf{content}:"/CFIDE/administrator"; http\_uri; nocase; \\
    iisadmin access\MyComment{web-application-attack} & \cite{nikto} & RCE & HTTP \MyComment{& Adicionar} & \textbf{content}:"/iisadmin"; nocase; http\_uri; \\
    JBoss JMX\MyComment{ Console Beanshell Exploit} & \cite{nikto} & RCE & HTTP \MyComment{& Remote Code Execution} & \textbf{content}:"/HtmlAdaptor"; nocase; http\_uri; \textbf{content}:"action=inspect"; nocase; http\_uri; \textbf{content}:"bean"; nocase; http\_uri; \textbf{content}:"name="; http\_uri; \\
    Script tag in URI\MyComment{Possible Cross Site Scripting Attempt} & \cite{nikto} & RCE & HTTP \MyComment{& Adicionar} & \textbf{content}:"</script>"; nocase; http\_uri; \\
    WeBid LFI \MyComment{Local File Inclusion} & \cite{nikto} & LFI & HTTP & \textbf{content}:"GET "; depth:4; \textbf{content}:"/cron.php?"; http\_uri; nocase; \textbf{content}:"include\_path="; http\_uri; nocase; \textbf{content}:"../"; \\
    %%WordPress Login & \cite{wrccdc} & PVE & HTTP & \textbf{content}:"log="; http\_client\_body; \textbf{content}:"\&pwd="; http\_client\_body; \textbf{content}:"\&wp-submit="; http\_client\_body; \\
    .htaccess access\MyComment{attempted-recon} & \cite{nikto} & AREC & HTTP \MyComment{& Adicionar} & \textbf{content}:".htaccess"; nocase; http\_uri; \\
    .idq access\MyComment{EXPLOIT ISAPI} & \cite{nikto} & RCE & HTTP \MyComment{& Adicionar} & \textbf{content}:".idq"; nocase; http\_uri; \\
    /system32/ in URI\MyComment{Possible Protected Directory Access Attempt} & \cite{nikto} & AREC & HTTP \MyComment{& Adicionar} & \textbf{content}:"/system32/"; nocase; http\_uri; \\
    Basebuilder LFI & \cite{nikto} & LFI & HTTP & \textbf{content}:"GET"; http\_method; \textbf{content}:"/main.inc.php?"; nocase; http\_uri; \textbf{content}:"mj\_config[src\_path]="; nocase; http\_uri; \\
    Tomcat server snoop access & \cite{nikto} & AREC & HTTP & \textbf{content}:"/jsp/snp/"; http\_uri; \textbf{content}:".snp"; http\_uri;\\  
    \bottomrule
  \end{tabular}
\end{table*}
